Exploiting Intrinsic Noise for Heterogeneous Cell Control Under Time Delays and Model Uncertainties

### 2023-08-21 paper.tex
update

Uncertain fluctuations, or `noise,' is a common theme throughout many fields of engineering, and is a frequent concern when attempting to control any human-made system such as vehicles \brian[\cite{XXX}], production machinery \brian[\cite{XXX}], or chemical processes \brian[\cite{XXX}].
\brian[Please look for some relevant references or a review of control under noise. I highlight several missing refs in red.]
Noise arrises from many sources, occasionally from quantum events and thermally-induced fluctuations, but more commonly from unknown or uncharacterized physical processes, and as such, these fluctuations usually cannot be efficiently reproduced or predicted using purely physical or mechanistic models.
Rather, statistical models are usually employed where key mechanisms or dynamics are subject to stochastic variations or inputs that act as a proxy for random or un-modeled fluctuations elsewhere in the system.
In this context, much effort has been placed on enhancing the robustness of controlled processes, where uncertain or unpredictable variations in internal or external parameters are modeled as intrinsic or extrinsic noise \brian[\cite{XXX}].
For many macroscopic, human-engineered systems, the resulting system dynamics can be modeled effectively simply by combining a deterministic (i.e., noise-free) model with additive noise (usually assumed to be Gaussian under arguments based on the central limit theorem) \brian[\cite{XXX}], and most current approaches seek to control the system to minimize variations that result from the noisy inputs.

Analyses of noise and control theory are equally relevant to understand the basic biological processes of gene transcription regulation \brian[\cite{XXX}] or mRNA translation regulation \brian[\cite{XXX}]or to modify these processes for practical use \brian[\cite{XXX}].
However, at this mesoscopic scale of cellular biology, where transcription factors compete to activate or deactivate individual genes or where mRNA or protein molecules are present at just a few copies (or none at all) per cell, the additive noise model is much less realistic.
In this case, Brownian motion, discrete stochastic gene regulation, and noisy mRNA dynamics collectively generate a fundamentally stochastic environment that cells must effectively manage. 
At this scale, the order or timing of a single reaction event (e.g., the binding of a transcription factor to a promoter) can have dramatic consequences that could last for several cellular generations (e.g., the activation or repression of a gene that promotes unfettered cell growth and differentiation). 
The cell's drive towards homeostasis requires dealing with the inherently chaotic and noisy processes that reside within it, and despite these challenges, cells generally demonstrate strong capability to survive these noisy processes. 
When seeking to understand how such systems evolve or react, the central limit theorem and the Gaussian noise may not apply, and a more detailed statistical analysis of probabilistic behavior is needed uncover hidden properties of cellular control mechanisms.

One emerging field that is particularly dependent on the integration of control and noise is synthetic biology, which aims to develop modular\cite{Ng2019} and orthogonal \cite{Liu2018} components to sense and manipulate \cite{Sheets2020} complex logical systems, enabling them to exhibit a wide range of advanced biological behaviors\cite{Shin2020}. 
Advances in optogenetics have enhanced the ability to reliably actuate embedded systems within cells, offering the potential to exert precise temporal and spatial control on cellular components\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}.
These developments have facilitated computer-programmable regulation of cellular protein production through external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}. 
These digital-synthetic actuators enable fine-tuned, computer-modulated control of cellular systems, previously unattainable, with faster response times compared to chemical diffusion\cite{Rullan2018, Baumschlager2017} . 
Classical and modern control methods like PID control and model predictive control have been implemented in such systems \brian[\cite{XXX}] to control synthetic systems to different stable points.

While most previous control efforts focussed on deterministic ordinary differential equation (ODE) formulations of the control problem, a few pioneering theoretical works have extended these analysis to include noise and have used stochastic simulations to show how systems can be guided towards unstable points by oscillating around those points \cite{Guarino2020}. 
It has recently been shown that new control techniques that leverage the complete probability distribution information of the system could actually harness the noise of single-cell gene regulation to achieve more complicated control objectives. 
For example, inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that noise could be exploited to achieve independent control of multiple cells using a single input, even despite uncertain parameters or time delays due to maturation of fluorescent proteins or limited observation of the regulatory proteins. 
In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al. \brian[\cite{XXX}] for the expression from a transcription promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure \ref{cartoons}A, top promoter).
We then proposed the addition of a positive auto-regulation (Figure \ref{cartoons}A, bottom promoter) to help maintain an elevated expression phenotype in the presence of UV excitation, and we demonstrated how a Single-Input-Multiple-Output (SIMO) multicellular controller could control multiple cells to arbitrary phenotypes using only a single input.

@@

\section{Methods}
In May et al. \cite{May2021}, we developed two models for the description of an optogenetically controlled gene expression system.
These first model consisted of six species to describe the light-activated association of two T7 split domains (species 1 and 2) which combine to form an active T7 polymerase (species 3) under optogenetic excitation. 
The active polymerase could then associate with inactive T7 promoters (species 4), resulting in the formation of an active allele (species 5) that could then transcribe and translated to produce the desired protein product (species 6). 
The second, much simpler, single-species model was developed by assuming quasi-steady equilibrium for the first five species. 
Both models were independently parameterized using the same experimental data from Baumschlager et al. \brian[\cite{XXX}]. 
Furthermore, an extension was made to each model to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}.
Through simulations, we showed that a feedback control law could be designed to force multiple cells to different and individually chosen equilibrium states using a single optogenetic control signal. 
We also showed that when this control law was parametrized using the simple model, it could be used effectively to control the behavior of the more complicated system, thus demonstrating that control performance could remain high despite inaccuracies in the model.
In the current work, we adopt the simpler of the two models and extend our analyses to consider the effects of additional model inaccuracies, including coarse-grained model approximations, parameter errors, and time delays, and we also explore the possibility that a single optogenetic control signal could drive the system to track temporally-changing reference signals.
Section \ref{sec:Model} introduces the model; Section \ref{sec:Stochastic} develops a Master Equation description of the models probabilistic dynamics; Section \ref{sec:Quantification} defines a control objective and optimizes that metric to obtain a baseline control law; and Sections \ref{sec:Scaling} and \ref{sec:Delay} introduce uncertainties into the model related to the granularity of the model approximation or the introduction of time delays, respectively.

@@

In realistic applications, models are never exact, but are often chosen as simplifications of known processes. 
For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection \brian[\cite{XXX}], time scale separations \brian[\cite{Peles2006}], Krylov subspaces \brian[\cite{Macnemara&Sidje?}], principle orthogonal decompositions \brian[\cite{Vo2019}], or other coarse meshes \brian[\cite{Munsky2008IEEE,Tapia2012CDC}]. 
Similarly, measurements are also always inexact and in many cases may only provide information about relative changes -- for example, although fluorescent proteins usually cannot be counted exactly, one may reasonably assume that a cell's total fluorescence intensity varies linearly with the fluorescent protein concentration. 
To explore how mismatches in the assumed system scale (e.g., arising from model approximations or relative measurements) affect the controllability of the cellular process, we define a granularity parameter ($\alpha=M'/M$) that linearly scales each species' population to increase ($\alpha>1$) or decrease ($\alpha<1$), while maintaining the dynamics and general behavior of the model. 
To apply this granularity parameter, we assume that each propensity function, $w_{\mu}$ (Eqns.\ \ref{Props}) is rescaled to a different level of discreteness by substituting
\begin{equation}\label{eq:granscale}
w'(\mathbf{X})=w(\mathbf{X}/\alpha).
\end{equation}
For example, the production and degradation of protein in cell one would become:
\begin{eqnarray}
w'_1(\mathbf{X}) &= u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0, \nonumber \\
 w'_2(\mathbf{X}) &= \gamma x_1/\alpha. \label{eq:rescaling}
\end{eqnarray}
We note that in order to reuse a control law that has been defined for one level granularity and apply it to a system at another level of granularity, the inputs to the controller must also be scaled by $1/\alpha$ before computing the  control level assigned to the current state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$. Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, control signal values at fractional state values after rescaling $(x_1/\alpha,x_2/\alpha)$ are calculated using 2D cubic interpolation from the control values at the nearest integer state values.
Finally, to provide a consistent metric for relative scoring, the definition of the performance score is also adjusted according to scale magnitudes. For example, in the two cell system the new steady state performance error would become:
\begin{eqnarray}
(J^2)' &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
& =\mathbf{C}'\mathbf{P}^*.
\label{EuclidV}
\end{eqnarray}
We reiterate that the system parameters and control law were defined and fixed using the the base granularity ($\alpha=1$), and to simulate a practical application where scales may be unknown or variable, these are not recomputed or refit upon changing the system granularity.

@@

\begin{equation}
u_{\tau}(t)=\left\{
\begin{array}{rl}
      0 ,&\textrm{ for }  t \leq \tau, \\
      {u}^{\mathcal{C}}(x_1(t-\tau), x_2(t-\tau)) , &\textrm{ for }   t > \tau,\\
\end{array}\right. 
\label{timeDelaySSA}
\end{equation}
where $\tau$ is the time delay between observation and actuation, ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$), and $x$ gathered from the state history. 
We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.
\brian[Still missing reference to the extrande method at this website
\url{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004923}].

\subsection{Tracking Time-Varying Trajectories}\label{sec:Dynamic}
Because optimizing the control law for a static set point as described in Section \ref{sec:Quantification} requires differentiation of the CME (Eq. \ref{CME}) with respect to the control signal at each state, this calculation is approaching the limits of current feasibility.
Extending these calculations to optimize controllers for a dynamically moving set point is much harder and would likely require intractable numerical simulation or development of new mathematical approaches that are beyond the scope of the current study. 
To circumvent this challenge, we instead propose a simple alternative in which the controller sweep though a piecewise constant set of controllers each designed for a specific static target point along the desired trajectory.

@@

The effect that granularity has on the control performance depends on two competing phenomena: First, because the controller was optimized for one level of granularity ($\alpha = 1$), one might expect that the controller would become worse if the granularity were incorrect. However, at large granularity, we find that the opposite is true - the control performance actually improves when applied to an incorrect model. The reason for this surprising result is that relative amplitude of stochastic fluctuations (i.e., the standard deviation divided by the mean) in a chemical process decreases with the inverse square root of the process scale \brian[\cite{VanKampen}]. In other words, the process becomes more predictable and therefore more controllable. At the extreme as the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that would separate different steady state behaviors \brian[\cite{XXX}].

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{DelayAndGranularity.pdf}
\vspace{-0.1in}
\caption{{\bf Joint effects of granularity and time delay on control performance.} (A-I) Joint probability distributions at different combinations of $\tau$ (rows) and $\alpha$ (columns). Overall RMSE shown at top. (J) Heat-map of control performance versus $(\tau,\alpha)$. Points A-I correspond to panels A-I. Dashed yellow line shows characteristic time $\tau_c = \alpha/\gamma/$.
\brian[Label 'J' is missing from the figure. Replace XXX below.]
(K) Controlled system trajectory for $\alpha = XXX$ and $\tau = $ min (denoted by red star in panel J).}
\label{DG}
\end{center}
\vspace{-0.2in}
\end{figure*}

@@

\begin{thebibliography}{10}
\providecommand{\url}[1]{#1}
\csname url@samestyle\endcsname
\providecommand{\newblock}{\relax}
\providecommand{\bibinfo}[2]{#2}
\providecommand{\BIBentrySTDinterwordspacing}{\spaceskip=0pt\relax}
\providecommand{\BIBentryALTinterwordstretchfactor}{4}
\providecommand{\BIBentryALTinterwordspacing}{\spaceskip=\fontdimen2\font plus
\BIBentryALTinterwordstretchfactor\fontdimen3\font minus
  \fontdimen4\font\relax}
\providecommand{\BIBforeignlanguage}[2]{{%
\expandafter\ifx\csname l@#1\endcsname\relax
\typeout{** WARNING: IEEEtran.bst: No hyphenation pattern has been}%
\typeout{** loaded for the language `#1'. Using the pattern for}%
\typeout{** the default language instead.}%
\else
\language=\csname l@#1\endcsname
\fi
#2}}
\providecommand{\BIBdecl}{\relax}
\BIBdecl

\bibitem{May2021}
M.~May and B.~Munsky, ``{Exploiting noise, nonlinearity, and feedback to
  differentially control multiple synthetic cells with a single optogenetic
  input},'' pp. 1--28, 2021.
eps
\bibitem{Ng2019}
A.~H. Ng, T.~H. Nguyen, M.~G{\'{o}}mez-Schiavon, G.~Dods, R.~A. Langan, S.~E.
  Boyken, J.~A. Samson, L.~M. Waldburger, J.~E. Dueber, D.~Baker, and
  H.~El-Samad, ``{Modular and tunable biological feedback control using a de
  novo protein switch},'' pp. 265--269, 2019.

\bibitem{Liu2018}
C.~C. Liu, M.~C. Jewett, J.~W. Chin, and C.~A. Voigt, ``{Toward an orthogonal
  central dogma},'' \emph{Nature Chemical Biology}, vol.~14, no.~2, pp.
  103--106, 2018.

\bibitem{Sheets2020}
M.~B. Sheets, W.~W. Wong, and M.~J. Dunlop, ``{Light-Inducible Recombinases for
  Bacterial Optogenetics},'' \emph{ACS Synthetic Biology}, vol.~9, no.~2, pp.
  227--235, 2020.

\bibitem{Groseclose2020}
\BIBentryALTinterwordspacing
T.~M. Groseclose, R.~E. Rondon, Z.~D. Herde, C.~A. Aldrete, and C.~J. Wilson,
  ``{Engineered systems of inducible anti-repressors for the next generation of
  biological programming},'' \emph{Nature Communications}, vol.~11, no.~1, pp.
  1--15, 2020. [Online]. Available:
  \url{http://dx.doi.org/10.1038/s41467-020-18302-1}
\BIBentrySTDinterwordspacing

\bibitem{Shin2020}
J.~Shin, S.~Zhang, B.~S. Der, A.~A. Nielsen, and C.~A. Voigt, ``{ Programming
  Escherichia coli to function as a digital display },'' \emph{Molecular
  Systems Biology}, vol.~16, no.~3, pp. 1--12, 2020.

\bibitem{Baumschlager2017}
A.~Baumschlager, S.~K. Aoki, and M.~Khammash, ``{Dynamic blue light-inducible
  T7 RNA polymerases (Opto-T7RNAPs) for precise spatiotemporal gene expression
  control},'' \emph{ACS synthetic biology}, vol.~6, no.~11, pp. 2157--2167,
  2017.

\bibitem{Chen2020}
\BIBentryALTinterwordspacing
S.~Y. Chen, L.~C. Osimiri, M.~Chevalier, L.~J. Bugaj, T.~H. Nguyen, R.~A.
  Greenstein, A.~H. Ng, J.~Stewart-Ornstein, L.~T. Neves, and H.~El-Samad,
  ``{Optogenetic Control Reveals Differential Promoter Interpretation of
  Transcription Factor Nuclear Translocation Dynamics},'' \emph{Cell Systems},
  vol.~11, no.~4, pp. 336----353.e24, 2020. [Online]. Available:
  \url{http://dx.doi.org/10.1016/j.cels.2020.08.009}
\BIBentrySTDinterwordspacing

\bibitem{Lillacci2018}
G.~Lillacci, Y.~Benenson, and M.~Khammash, ``{Synthetic control systems for
  high performance gene expression in mammalian cells},'' \emph{Nucleic acids
  research}, vol.~46, no.~18, pp. 9855--9863, 2018.

\bibitem{Fox2021}
\BIBentryALTinterwordspacing
Z.~R. Fox, S.~Fletcher, A.~Fraisse, C.~Aditya, and S.~Sosa, ``{MicroMator: Open
  and Flexible Software for Reactive Microscopy},'' \emph{bioRxiv}, pp. 1--9,
  2021. [Online]. Available:
  \url{http://biorxiv.org/cgi/content/short/2021.03.12.435206v1}
\BIBentrySTDinterwordspacing

\bibitem{Baumschlager2021}
\BIBentryALTinterwordspacing
A.~Baumschlager and M.~Khammash, ``{Synthetic Biological Approaches for
  Optogenetics and Tools for Transcriptional Light-Control in Bacteria},''
  \emph{Advanced Biology}, vol.~5, no.~5, p. 2000256, 2021. [Online].
  Available:
  \url{https://onlinelibrary.wiley.com/doi/abs/10.1002/adbi.202000256}
\BIBentrySTDinterwordspacing

\bibitem{Lugagne2017}
\BIBentryALTinterwordspacing
J.~B. Lugagne, S.~{Sosa Carrillo}, M.~Kirch, A.~K{\"{o}}hler, G.~Batt, and
  P.~Hersen, ``{Balancing a genetic toggle switch by real-time feedback control
  and periodic forcing},'' \emph{Nature Communications}, vol.~8, no.~1, pp.
  1--7, 2017. [Online]. Available:
  \url{http://dx.doi.org/10.1038/s41467-017-01498-0}
\BIBentrySTDinterwordspacing

\bibitem{Rullan2018}
\BIBentryALTinterwordspacing
M.~Rullan, D.~Benzinger, G.~W. Schmidt, A.~Milias-Argeitis, and M.~Khammash,
  ``{An Optogenetic Platform for Real-Time, Single-Cell Interrogation of
  Stochastic Transcriptional Regulation},'' \emph{Molecular Cell}, vol.~70,
  no.~4, pp. 745--756.e6, 2018. [Online]. Available:
  \url{https://doi.org/10.1016/j.molcel.2018.04.012}
\BIBentrySTDinterwordspacing

\bibitem{Guarino2020}
A.~Guarino, D.~Fiore, D.~Salzano, and M.~{Di Bernardo}, ``{Balancing Cell
  Populations Endowed with a Synthetic Toggle Switch via Adaptive Pulsatile
  Feedback Control},'' \emph{ACS Synthetic Biology}, vol.~9, no.~4, pp.
  793--803, 2020.

\bibitem{Munsky2012}
B.~Munsky, G.~Neuert, and A.~Van~Oudenaarden, ``{Using gene expression noise to
  understand gene regulation},'' \emph{Science}, vol. 336, no. 6078, pp.
  183--187, 2012.

\bibitem{Szymanska2015}
P.~Szyma{\'{n}}ska, N.~Gritti, J.~M. Keegstra, M.~Soltani, and B.~Munsky,
  ``{Using noise to control heterogeneity of isogenic populations in homogenous
  environments},'' \emph{Physical Biology}, vol.~12, no.~4, 2015.

\bibitem{Gillespie1992}
D.~T. Gillespie, ``{A rigorous derivation of the chemical master equation},''
  \emph{Physica A: Statistical Mechanics and its Applications}, vol. 188, no.
  1-3, pp. 404--425, 1992.

\bibitem{Gillespie1977}
D.~T. Gillespie, ``{Exact stochastic simulation of coupled chemical reactions},''
  \emph{The journal of physical chemistry}, vol.~81, no.~25, pp. 2340--2361,
  1977.

\end{thebibliography}

\end{document}

>>>>

Uncertain fluctuations, or `noise,' is a common theme throughout many fields of engineering, and robust control is a frequent concern when attempting to control any human-made system such as vehicles \michael[\cite{Liang2021}], chemical processes \michael[\cite{Lucia2013}], or biology \michael[\cite{Baetica2020}].
\brian[Please look for some relevant references or a review of control under noise. I highlight several missing refs in red.]
Noise arrises from many sources, occasionally from quantum events and thermally-induced fluctuations, but more commonly from unknown or uncharacterized physical processes, and as such, these fluctuations usually cannot be efficiently reproduced or predicted using purely physical or mechanistic models.
Rather, statistical models are usually employed where key mechanisms or dynamics are subject to stochastic variations or inputs that act as a proxy for random or un-modeled fluctuations elsewhere in the system.
In this context, much effort has been placed on enhancing the robustness of controlled processes, where uncertain or unpredictable variations in internal or external parameters are modeled as intrinsic or extrinsic noise \brian[cite XXX].
For many macroscopic, human-engineered systems, the resulting system dynamics can be modeled effectively simply by combining a deterministic (i.e., noise-free) model with additive noise (usually assumed to be Gaussian under arguments based on the central limit theorem) \brian[\cite{Feng2022}], and most current approaches seek to control the system to minimize variations that result from the noisy inputs.

Analyses of noise and control theory are equally relevant to understand the basic biological processes of gene transcription regulation \michael[\cite{Sun2020}] or mRNA translation regulation \brian[\cite{Cialek2022}]or to modify these processes for practical use \brian[\cite{Dionisi2022,Baumschlager2017}].
However, at this mesoscopic scale of cellular biology, where transcription factors compete to activate or deactivate individual genes or where mRNA or protein molecules are present at just a few copies (or none at all) per cell, the additive noise model is much less realistic.
In this case, Brownian motion, discrete stochastic gene regulation, and noisy mRNA dynamics collectively generate a fundamentally stochastic environment that cells must effectively manage. 
At this scale, the order or timing of a single reaction event (e.g., the binding of a transcription factor to a promoter) can have dramatic consequences that could last for several cellular generations (e.g., the activation or repression of a gene that promotes unfettered cell growth and differentiation). 
The cell's drive towards homeostasis requires dealing with the inherently chaotic and noisy processes that reside within it, and despite these challenges, cells generally demonstrate strong capability to survive these noisy processes. 
When seeking to understand how such systems evolve or react, the central limit theorem and the Gaussian noise may not apply, and a more detailed statistical analysis of probabilistic behavior is needed uncover hidden properties of cellular control mechanisms.

One emerging field that is particularly dependent on the integration of control and noise is synthetic biology, which aims to develop modular\cite{Ng2019} and orthogonal \cite{Liu2018} components to sense and manipulate \cite{Sheets2020} complex logical systems, enabling them to exhibit a wide range of advanced biological behaviors\cite{Shin2020}. 
Advances in optogenetics have enhanced the ability to reliably actuate embedded systems within cells, offering the potential to exert precise temporal and spatial control on cellular components\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}.
These developments have facilitated computer-programmable regulation of cellular protein production through external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}. 
These digital-synthetic actuators enable fine-tuned, computer-modulated control of cellular systems, previously unattainable, with faster response times compared to chemical diffusion\cite{Rullan2018, Baumschlager2017} . 
Classical and modern control methods like PID control and model predictive control have been implemented in such systems \brian[\cite{Filo2022}] to control synthetic systems to different stable points.

It has recently been shown that new control techniques that leverage the complete probability distribution information of the system could actually harness the noise of single-cell gene regulation to achieve more complicated control objectives. 
For example, inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that noise could be exploited to achieve independent control of multiple cells using a single input, even despite uncertain parameters or time delays due to maturation of fluorescent proteins or limited observation of the regulatory proteins. 
In May et al.~\cite{May2021}, we identified a simplified stochastic model to reproduce data measured in Baumschlager, et al. \brian[\cite{Baumschlager2017}] for the expression from a transcription promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure \ref{cartoons}A, top promoter).
We then proposed the addition of a positive auto-regulation (Figure \ref{cartoons}A, bottom promoter) to help maintain an elevated expression phenotype in the presence of UV excitation, and we demonstrated how a Single-Input-Multiple-Output (SIMO) multicellular controller could control multiple cells to arbitrary phenotypes using only a single input.

@@

\section{Methods}
In May et al. \cite{May2021}, we developed two models for the description of an optogenetically controlled gene expression system.
These first model consisted of six species to describe the light-activated association of two T7 split domains (species 1 and 2) which combine to form an active T7 polymerase (species 3) under optogenetic excitation. 
The active polymerase could then associate with inactive T7 promoters (species 4), resulting in the formation of an active allele (species 5) that could then transcribe and translated to produce the desired protein product (species 6). 
The second, much simpler, single-species model was developed by assuming quasi-steady equilibrium for the first five species. 
Both models were independently parameterized using the same experimental data from Baumschlager et al. \brian[\cite{Baumschlager2017}]. 
Furthermore, an extension was made to each model to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}.
Through simulations, we showed that a feedback control law could be designed to force multiple cells to different and individually chosen equilibrium states using a single optogenetic control signal. 
We also showed that when this control law was parametrized using the simple model, it could be used effectively to control the behavior of the more complicated system, thus demonstrating that control performance could remain high despite inaccuracies in the model.
In the current work, we adopt the simpler of the two models and extend our analyses to consider the effects of additional model inaccuracies, including coarse-grained model approximations, parameter errors, and time delays, and we also explore the possibility that a single optogenetic control signal could drive the system to track temporally-changing reference signals.
Section \ref{sec:Model} introduces the model; Section \ref{sec:Stochastic} develops a Master Equation description of the models probabilistic dynamics; Section \ref{sec:Quantification} defines a control objective and optimizes that metric to obtain a baseline control law; and Sections \ref{sec:Scaling} and \ref{sec:Delay} introduce uncertainties into the model related to the granularity of the model approximation or the introduction of time delays, respectively.

@@

In realistic applications, models are never exact, but are often chosen as simplifications of known processes. 
For example, when analyzing discrete stochastic chemical kinetics, it is common to project the CME onto lower-dimensional spaces using finite state projection \brian[\cite{Munsky2006}], time scale separations \brian[\cite{Peles2006}], Krylov subspaces \brian[\cite{Macnemara2008}], principle orthogonal decompositions \brian[\cite{Vo2019}], or other coarse meshes \brian[\cite{Munsky2008IEEE,Tapia2012CDC}]. 
Similarly, measurements are also always inexact and in many cases may only provide information about relative changes -- for example, although fluorescent proteins usually cannot be counted exactly, one may reasonably assume that a cell's total fluorescence intensity varies linearly with the fluorescent protein concentration. 
To explore how mismatches in the assumed system scale (e.g., arising from model approximations or relative measurements) affect the controllability of the cellular process, we define a granularity parameter ($\alpha=M'/M$) that linearly scales each species' population to increase ($\alpha>1$) or decrease ($\alpha<1$), while maintaining the dynamics and general behavior of the model. 
To apply this granularity parameter, we assume that each propensity function, $w_{\mu}$ (Eqns.\ \ref{Props}) is rescaled to a different level of discreteness by substituting
\begin{equation}\label{eq:granscale}
w'(\mathbf{X})=w(\mathbf{X}/\alpha).
\end{equation}
For example, the production and degradation of protein in cell one would become:
\begin{eqnarray}
w'_1(\mathbf{X}) &= u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0, \nonumber \\
 w'_2(\mathbf{X}) &= \gamma x_1/\alpha. \label{eq:rescaling}
\end{eqnarray}
We note that in order to reuse a control law that has been defined for one level granularity and apply it to a system at another level of granularity, the inputs to the controller must also be scaled by $1/\alpha$ before computing the  control level assigned to the current state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$. Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, control signal values at fractional state values after rescaling $(x_1/\alpha,x_2/\alpha)$ are calculated using 2D cubic interpolation from the control values at the nearest integer state values.
Finally, to provide a consistent metric for relative scoring, the definition of the performance score is also adjusted according to scale magnitudes. For example, in the two cell system the new steady state performance error would become:
\begin{eqnarray}
(J^2)' &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
& =\mathbf{C}'\mathbf{P}^*.
\label{EuclidV}
\end{eqnarray}
We reiterate that the system parameters and control law were defined and fixed using the the base granularity ($\alpha=1$), and to simulate a practical application where scales may be unknown or variable, these are not recomputed or refit upon changing the system granularity.

@@

\begin{equation}
u_{\tau}(t)=\left\{
\begin{array}{rl}
      0 ,&\textrm{ for }  t \leq \tau, \\
      {u}^{\mathcal{C}}(x_1(t-\tau), x_2(t-\tau)) , &\textrm{ for }   t > \tau,\\
\end{array}\right. 
\label{timeDelaySSA}
\end{equation}
where $\tau$ is the time delay between observation and actuation, ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$), and $x$ gathered from the state history. 
We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.
\brian[Still missing reference to the extrande method at this website
\url{https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004923}].
\michael[added Voliotis2016 to the section after this comment]
The Extrande method \cite{Voliotis2016} was used to update the control input at an average frequency of 50 updates per minute, far exceeding the dynamics of the system.
\subsection{Tracking Time-Varying Trajectories}\label{sec:Dynamic}
Because optimizing the control law for a static set point as described in Section \ref{sec:Quantification} requires differentiation of the CME (Eq. \ref{CME}) with respect to the control signal at each state, this calculation is approaching the limits of current feasibility.
Extending these calculations to optimize controllers for a dynamically moving set point is much harder and would likely require intractable numerical simulation or development of new mathematical approaches that are beyond the scope of the current study. 
To circumvent this challenge, we instead propose a simple alternative in which the controller sweep though a piecewise constant set of controllers each designed for a specific static target point along the desired trajectory.

@@

The effect that granularity has on the control performance depends on two competing phenomena: First, because the controller was optimized for one level of granularity ($\alpha = 1$), one might expect that the controller would become worse if the granularity were incorrect. However, at large granularity, we find that the opposite is true - the control performance actually improves when applied to an incorrect model. The reason for this surprising result is that relative amplitude of stochastic fluctuations (i.e., the standard deviation divided by the mean) in a chemical process decreases with the inverse square root of the process scale \brian[\cite{VanKampen}]. In other words, the process becomes more predictable and therefore more controllable. At the extreme as the system size increases, the dynamics converge towards a deterministic process, except for certain exceptional initial conditions lying on manifolds that would separate different steady state behaviors \brian[\cite{Strasser2012}].

@@

\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{DelayAndGranularity.pdf}
\vspace{-0.1in}
\caption{{\bf Joint effects of granularity and time delay on control performance.} (A-I) Joint probability distributions at different combinations of $\tau$ (rows) and $\alpha$ (columns). Overall RMSE shown at top. (J) Heat-map of control performance versus $(\tau,\alpha)$. Points A-I correspond to panels A-I. Dashed yellow line shows characteristic time $\tau_c = \alpha/\gamma/$.
\brian[Label 'J' is missing from the figure. Replace XXX below.] \michael[added J. Replace XXX with 4.0]
(K) Controlled system trajectory for $\alpha = 4.0$ and $\tau = $ min (denoted by red star in panel J).}
\label{DG}
\end{center}
\vspace{-0.2in}
\end{figure*}

@@

\section*{References}
\bibliographystyle{plain}
\bibliography{refs}
\end{document}
